Increasing productivity amid stable nutrient regmines in Rhode Island Lakes and Reservoirs

### 2019-11-08 manuscript/manuscript.Rmd
more slides and lagosne available data stats added to script (not yet to ms)

Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).   URIWW is a scientist-led citizen science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.  Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g. May through October).  The entire effort follow rigorous quality control/quality assurance protocols.  These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing, @Oliver_2017].  URIWW data contributed to the larger regional study by Oliver et al. [-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.

>>>>

Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).   URIWW is a scientist-led citizen science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.  Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g. May through October).  The entire effort follow rigorous quality control/quality assurance protocols.  These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing; @Oliver_2017].  URIWW data contributed to the larger regional study by Oliver et al. [-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.

### 2019-11-08 nalms_2019_talk/increasing_chla_stable_nutrients_talk.Rmd
more stuff in slides
more slides and lagosne available data stats added to script (not yet to ms)

- Parameters: Total Nitrogen, Total Phosphorus, N:P, Temperature, Chlorophyll
- Years: 1993 to 2016
- Months: May through October
- Depths: <= 2 meters
- Sites: At least 10 years of data
- Total of 69 Sites

@@

## LAGOSNE

@@

@@

## Results: Nutrient trends

@@

@@

>>>>

- Parameters: Total Nitrogen, Total Phosphorus, N:P, Temperature, Chlorophyll
- Years: 1993 to 2016
- Months: May through October
- Depths: <= 2 meters
- Sites: At least 10 years of data
- Total of 69 Sites
  - 66 with all parameters
  - 1 without N:P
  - 2 without chlorophyll and temperature

@@

## LAke mult-scaled GeOSpatial and temoporal database for the Northeastern US (LAGOS-NE)

- Soranno et al (2017). LAGOS-NE: a multi-scaled geospatial and temporal database of lake ecological context and water quality for thousands of US lakes. Gigascience, 6(12)
- Stachelek and Oliver (2017). LAGOSNE: Interface to the lake multi-scaled geospatial and temporal database. R package version 1.1.0. https://cran.r-project.org/package=LAGOSNE
- <https://lagoslakes.org/>
- 17 States: CT, IL, IN, IA, ME, MA, MI, MN, MO, NH, NJ, NY, OH, PA, RI, VT, WI
- ~ 50,000 Lakes

![](figures/gix101fig1.jpeg)

## LAke mult-scaled GeOSpatial and temoporal database for the Northeastern US (LAGOS-NE): Filtered for Trend Analysis

- Parameters: Total Nitrogen, Total Phosphorus, N:P, Chlorophyll
  - Temperature data not currently available
- Years: 1993 to 2016
- Months: May through October
- Depths: <= 2 meters
- Sites: At least 10 years of data
- Total of 1482 Sites
  - 637 with chlorophyll and phosphorus
  - 377 with only chlorophyll
  - 275 with only phosphorus
  - 161 with all variables (except temperature)
  - 25 with chlorophyll and nitrogen
  - 7 with other combination of nutrients

@@

![](figures/z-score_schematic.jpg)

@@

## Results: Nitrogen trends

![](figures/ww_tn_trends.jpg)

![](figures/lagos_tn_trends.jpg)

## Results: Phosphorus trends

![](figures/ww_tp_trends.jpg)

![](figures/lagos_tp_trends.jpg)

## Results: N:P trends

![](figures/ww_np_trends.jpg)

![](figures/lagos_np_trends.jpg)

@@

![](figures/ww_chla_trends.jpg)

![](figures/lagos_chla_trends.jpg)

@@

![](figures/ww_temp_trends.jpg)
